---
title: "Lab 2 - Computational Statistics"
author: "Gustav Sternelöv"
date: "8 februari 2016"
output: pdf_document
---

# Assignment 1

## 1.1-1.2
Data is imported and splitted into a training set and a test set. 
```{r, echo=FALSE}
mort_rate<-read.csv2("C:\\Users\\Gustav\\Documents\\Computational-Statistics\\Lab2\\mortality_rate.csv",sep=";")
mort_rate$LMR <- log(mort_rate$Rate)

n=dim(mort_rate)[1]
set.seed(123456)
id=sample(1:n, floor(n*0.5))
train=mort_rate[id,]
test=mort_rate[-id,]
```
Thereafter, the function *myMSE* is written. It takes the arguments pars (a list including *X*, *Y*, *XTest* and *YTest*) and the argument *lambda*. The latter parameter specifies the penalty parameter *enp.target* in *loess*. The *myMSE* function then uses *loess* to make a model which is evaluated on the test set. Returned by *myMSE* is the predictive MSE. 
```{r}
myMSE <- function(lambda, pars){
  model <- loess(pars$Y~pars$X, data=pars[1:2], enp.target = lambda)
  Pred <- predict(model, newdata=pars$Xtest)
  MSE <- (1/length(pars$Y)) * sum((Pred-pars$Ytest)^2)
  print(MSE)
  return(MSE)
}
```

## 1.3
```{r, echo=FALSE, results='hide'}
Data <- list(X=train$Day, Y=train$LMR, Xtest=test$Day, Ytest=test$LMR)
# 1.3 
Lambda <-  seq(0.1,40, by=0.1)
modelMSE <- 0
j <- 0
for(i in Lambda){
  j <- j+1
  modelMSE[j] <- myMSE(Lambda[j], Data)
}
```

```{r, echo=FALSE, fig.height=3.7}
library(ggplot2)
MSEdata <- data.frame(MSE=modelMSE, lambda=Lambda)  
ggplot(MSEdata, aes(x=lambda, y=MSE, label=lambda)) + geom_line(col="darkblue")+
  geom_text(data=subset(MSEdata, MSE <= min(MSEdata$MSE))[1,],vjust=-0.6, size=10) + 
  geom_point(data=subset(MSEdata, MSE <= min(MSEdata$MSE))[1,], size=5, col="darkorange")
```

## 1.4

```{r, echo=FALSE}
Optim <- optimize(f=myMSE, interval = c(0.1, 40), tol=0.01, pars=Data)
Optim
```


## 1.5

```{r, echo=FALSE}
optim(par=35, fn=myMSE, method = "BFGS", pars=Data)
```


# Assignment 2

## 2.1-2.2

```{r, echo=FALSE}
# 2.1
load("C:\\Users\\Gustav\\Documents\\Computational-Statistics\\Lab2\\data.RData")
Data2 <- data
# Maximum likelihood estimates
# my
MeanEst <- sum(Data2) / length(Data2)

# sigma2 
(1/length(Data2)) * sum((Data2-MeanEst)^2)

```


## 2.3


## 2.4
